---
title: "Final Project"
output: html_document
date: "2025-03-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading packages

```{r}
library(tidyverse)
library(ggplot2)
library(sf)
library(tmap)
library(SpatialKDE)
library(terra)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)
library(classInt)
library(spdep)
library(raster)
library(mapview)
```

#Connecting datasets

```{r}
ev.stations <- read_csv("./data/ev charging station.csv")
ev.users <- read_csv("./data/ev registration counts.csv")
gdp <- read.csv("./data/GDP_edited.csv")
veh.reg <- read.csv("./data/Vehicle registration.csv")

state <- read_sf("./data/state/state.shp")
evcs <- read_sf("./data/evcs.gpx")
#midwest <- vect("./data/state/midwest.shp")
#landcover <- terra::rast("./data/Annual_NLCD_LndCov_2023_CU_C1V0.tif")
```

#Exploring data

```{r}
view(ev.stations)
view(ev.users)
view(state)
```

#Cleaning/Organizing data

```{r}
#removing unnecessary states from datasets
ev.stations <- ev.stations %>% dplyr::filter(State!="AK"& State!="HI"& State!="PR")
ev.users <- ev.users %>% dplyr::filter(State_abb!="AK"& State_abb!="HI"& State_abb!="Total")
gdp <- gdp %>% dplyr::filter(State!="Alaska"& State!="Hawaii")

```

#Road file organizing for midwest zone********************************************
#Preparing the road dataset to work but couldn't load due to file size and environment storage issue.
***No need to run again***

```{r}
# oh <- sf::read_sf("./data/Trans/TRAN_Ohio_State_GPKG/TRAN_Ohio_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# il <- sf::read_sf("./data/Trans/TRAN_Illinois_State_GPKG/TRAN_Illinois_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# In <- sf::read_sf("./data/Trans/TRAN_Indiana_State_GPKG/TRAN_Indiana_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# ws <- sf::read_sf("./data/Trans/TRAN_Wisconsin_State_GPKG/TRAN_Wisconsin_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# iw <- sf::read_sf("./data/Trans/TRAN_Iowa_State_GPKG/TRAN_Iowa_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# mc <- sf::read_sf("./data/Trans/TRAN_Michigan_State_GPKG/TRAN_Michigan_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# nb <- sf::read_sf("./data/Trans/TRAN_Nebraska_State_GPKG/TRAN_Nebraska_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# kn <- sf::read_sf("./data/Trans/TRAN_Kansas_State_GPKG/TRAN_Kansas_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# ms <- sf::read_sf("./data/Trans/TRAN_Missouri_State_GPKG/TRAN_Missouri_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# mn <- sf::read_sf("./data/Trans/TRAN_Minnesota_State_GPKG/TRAN_Minnesota_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# sd <- sf::read_sf("./data/Trans/TRAN_South_Dakota_State_GPKG/TRAN_South_Dakota_State_GPKG.gpkg", layer = "Trans_RoadSegment")
# nd <- sf::read_sf("./data/Trans/TRAN_North_Dakota_State_GPKG/TRAN_North_Dakota_State_GPKG.gpkg", layer = "Trans_RoadSegment")
 
# merged_midwest <- rbind(oh, il, In, ws, mn, ms, mc, iw, nb, kn, sd, nd)
# st_write(merged_midwest, "./data/Trans/midwest_road.shp", overwrite=TRUE)

```

#Descriptive Analysis

```{r}
#CS access types (public and private) with additional details*********************
ev.access <- ev.stations %>% as_tibble() %>% group_by(`Groups With Access Code`) %>% summarize(TypeCount = n())
write.csv(ev.access, "./result/ev.access.csv")

#CS Service locations with access type********************************************
access.facility <- ev.stations %>% as_tibble() %>% group_by(`Facility Type`) %>% summarize(TypeCount = n())
write.csv(access.facility, "./result/access.facility.csv")

access.facility2 <- ev.stations %>% as_tibble() %>% group_by(`Facility Type`,`Access Code`) %>% summarize(TypeCount = n(), .groups = "drop")
write.csv(access.facility2, "./result/access.facility2.csv")

ggplot(access.facility2, aes(x = `Facility Type`, y= TypeCount, fill=`Access Code`)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "EVCS locations", y = "Count", x = "Location")+
  theme_minimal() +
  theme (axis.text.x = element_text(angle=45, hjust=1)) 

#time frame analysis of establishment of EVCS************************************
ev.stations <- ev.stations %>% mutate(.,Open_yr = stringr::str_sub(`Open Date`, 1, 4))

ev.stations %>% group_by(Open_yr) %>% summarize(TypeCount = n()) %>%
ggplot(., aes(x = Open_yr, y = TypeCount, group=1)) +
  geom_line(color="orange", size=1.5)+
  geom_point()+
  labs(title = "EVCS Placement From Year 1995 to 2025", y = "Count", x = "Opening Year")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle=45, hjust=1))

evcs.open <- ev.stations %>% group_by(Open_yr, `Access Code`) %>% summarize(TypeCount = n(), .groups = "drop")
write.csv(evcs.open, "./result/evcs.open.csv")

ggplot(evcs.open, aes(x = Open_yr, y= TypeCount, fill=`Access Code`)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "EVCS Placement year-wise", y = "Count", x = "Year")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle=45, hjust=1))

#state-wise CS of different accessibility*****************************************
ev.stations %>% as_tibble() %>% group_by(State,`Access Code`) %>% summarize(TypeCount = n(),.groups="drop") %>%
ggplot(., aes(x = State, y= TypeCount, fill=`Access Code`)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "State-wise Access Type", y = "Count", x = "State")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle=45, hjust=1))

#state-wise different types of Chargers*******************************************
ev.stations %>% group_by(State) %>% summarise(.,TotalCount = n(), Level1=sum(`EV Level1 EVSE Num`, na.rm = TRUE), Level2= sum(`EV Level2 EVSE Num`, na.rm = TRUE), Fast=sum(`EV DC Fast Count`, na.rm = TRUE), .groups = "drop")

charger.type <- ev.stations %>% dplyr::select(State, `Access Code`, `EV Level1 EVSE Num`,`EV Level2 EVSE Num`, `EV DC Fast Count`) %>% group_by(State, `Access Code`) %>% summarise(.,Level1=sum(`EV Level1 EVSE Num`, na.rm = TRUE), Level2= sum(`EV Level2 EVSE Num`, na.rm = TRUE), Fast=sum(`EV DC Fast Count`, na.rm = TRUE), .groups = "drop")
write.csv(charger.type, "./result/charger.type.csv")

#Chart of total count of different charger type
pivot_longer(charger.type,
             cols = Level1:Fast,
             names_to = "Type",
             values_to = "Value") %>%
  group_by(Type) %>% summarise(TotCount=sum(Value)) %>%
  ggplot(., aes(x = Type, y=TotCount)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Different Charger Frequency in the US",
       x = "Charger Type",
       y = "Count")+
  theme_minimal()

#Chart of total count of access type with charger type
pivot_longer(charger.type,
            cols = Level1:Fast,
            names_to = "Type",
            values_to = "Value") %>%
  group_by(Type, `Access Code`) %>% summarise(TotCount=sum(Value), .groups = "drop") %>%
  ggplot(., aes(x = Type, y=TotCount, fill =`Access Code`)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Different Chargertype Frequency with Accesstype in the US",
        x = "Charger Type",
        y = "Count")+
    theme_classic()

```

#Generating Maps (Static): comparing the distribution of EVCS with the demography

```{r}
#State map exploring
#basic state map
# ggplot(data = state) +
#   geom_sf() +
#   theme_minimal()

#state-wise total CS
state_cs <- ev.stations %>% as_tibble() %>% group_by(State) %>% summarise(StationsNo = n())
#Adding state-wise stations and EV users number with state map
state.users <- left_join(state, ev.users, by = c("STATE_NAME" = "State"))
state.cs <- left_join(state.users, state_cs, by = c("State_abb" = "State"))
state.cs <- state.cs %>% mutate(CsUserRatio=StationsNo/`Registration Count`)
state.cs <- state.cs %>% mutate(Pop_dens=POPULATION/AREASQKM)
state.cs <- left_join(state.cs, gdp, by = c("State_abb" = "State_Abb"))
write.csv(state.cs, "./result/state.cs.csv")

#symbology map with population density********************************************
tm_shape(state.cs) + 
  tm_polygons(fill = "Pop_dens", fill.scale = tm_scale_intervals(style = "quantile", values = "Blues")) +
  tm_layout(
    title = "Population distribution",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE,
  )

#symbology map with area**********************************************************
tm_shape(state.cs) + 
  tm_polygons(fill = "AREASQKM", fill.scale = tm_scale_intervals(style = "quantile", values = "Blues")) +
  tm_layout(
    title = "State Area (sq.km)",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE,
  )

#symbology map with EVCS**********************************************************
tm_shape(state.cs) + 
  tm_polygons(fill = "StationsNo", fill.scale = tm_scale_intervals(style = "quantile", values = "Blues")) +
  tm_layout(
    title = "EVCS distribution",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE
  )

#symbology map with EV users******************************************************
 tm_shape(state.cs) + 
   tm_polygons(fill = "Registration Count", fill.scale = tm_scale_intervals(style = "quantile", values = "Blues")) +
  tm_layout(
    title = "EV Users distribution",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE
  )

#symbology map with the ratio of cs to EV users***********************************
tm_shape(state.cs) + 
  tm_polygons(fill = "CsUserRatio", fill.scale = tm_scale_intervals(style = "quantile", values = "Blues")) +
  tm_layout(
    title = "User ratio with EVCS",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE
  ) +
  tm_labels("State_abb", size = 0.5)

#symbology map with state-wise gdp contribution***********************************
tm_shape(state.cs) + 
  tm_polygons(fill = "Millionsdollars_2024", fill.scale = tm_scale_intervals(style = "equal", values = "Blues")) +
  tm_layout(
    title = "State-wise GDP Contribution",
    title.size = 1,
    title.position = c("right", "top"),
    legend.position = c("right", "bottom"),
    legend.title.size = 0.8,
    legend.title.color = "black",
    legend.text.size = 0.7, 
    legend.frame = FALSE
  ) +
  tm_labels("State_abb", size = 0.5)
 
```

##Generating Maps (Dynamic): EVCS Placement time series visualization

```{r}
#EVCS map exploring
ggplot(data = evcs) +
   geom_sf() +
   theme_minimal()

points_cs <- st_as_sf(ev.stations, coords = c("Longitude", "Latitude"), crs = 4326) 
yr95_yr00 <- points_cs %>% dplyr::filter(Open_yr=="1995" | Open_yr=="1996" | Open_yr=="1997" | Open_yr=="1998" | Open_yr=="1999"| Open_yr=="2000")
yr01_yr05 <- points_cs %>% dplyr::filter(Open_yr=="2001" | Open_yr=="2002" | Open_yr=="2003" | Open_yr=="2004" | Open_yr=="2005")
yr06_yr10 <- points_cs %>% dplyr::filter(Open_yr=="2006" | Open_yr=="2007" | Open_yr=="2008" | Open_yr=="2009" | Open_yr=="2010")
yr11_yr15 <- points_cs %>% dplyr::filter(Open_yr=="2011" | Open_yr=="2012" | Open_yr=="2013" | Open_yr=="2014" | Open_yr=="2015")
yr16_yr20 <- points_cs %>% dplyr::filter(Open_yr=="2016" | Open_yr=="2017" | Open_yr=="2018" | Open_yr=="2019" | Open_yr=="2020")
yr21_yr25 <- points_cs %>% dplyr::filter(Open_yr=="2021" | Open_yr=="2022" | Open_yr=="2023" | Open_yr=="2024" | Open_yr=="2025")

leaflet(data = state.cs) %>% 
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB") %>% 
  addProviderTiles(providers$OpenStreetMap, group = "OpenStreetMap") %>% 
  addProviderTiles(providers$Esri.WorldImagery, group = "Esri Satellite") %>% 
  addPolygons( fillColor = "gray", stroke = TRUE, color = "black", weight = 0.3, popup = ~paste0(State), group="State") %>%
  addCircles(data=yr95_yr00, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "red", group="1995-2000") %>%
  addCircles(data=yr01_yr05, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "yellow", group="2001-2005") %>%
  addCircles(data=yr06_yr10, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "purple", group="2006-2010") %>%
  addCircles(data=yr11_yr15, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "green", group="2011-2015") %>%
  addCircles(data=yr16_yr20, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "orange", group="2016-2020") %>%
  addCircles(data=yr21_yr25, lng = ~st_coordinates(geometry)[,1],lat = ~st_coordinates(geometry)[,2],radius = 0.5, color = "blue", group="2021-2025") %>%
  addLayersControl(baseGroups = c("CartoDB", "OpenStreetMap", "Esri Satellite"), overlayGroups = c("State", "1995-2000","2001-2005", "2006-2010", "2011-2015","2016-2020","2021-2025"), 
                   position = "bottomleft", options = layersControlOptions(collapsed = FALSE)) %>%
  addLegend(position = "topright",
            colors = c("red", "yellow", "purple", "green", "orange", "blue"),
            labels = c("95-00", "01-05", "06-10","11-15", "16-20", "21-25"),
            title = "Placement Year") %>%
  addMiniMap()

```

#Kernal Density Model
#taking too much time to run, sometimes require to change the cell size and band_width
#This portion is not working properly*********************************************

```{r}
cell_size <- 50000
band_width <- 100000

state_proj <- st_transform(state.cs, crs = 32615)
points_cs_proj <- st_transform(points_cs, crs = 32615) #Had to change the projection because of error

grid_state <- st_make_grid(state_proj, cellsize = cell_size, offset = st_bbox(state_proj)[c("xmin", "ymin")] - band_width,
  square = TRUE)
grid_state <- st_sf(geometry = grid_state)
tm_shape(grid_state) + 
  tm_polygons()

kde <- points_cs_proj %>% SpatialKDE::kde(band_width = band_width, kernel = "quartic", grid = grid_state)
tm_shape(kde) +
  tm_polygons(col = "kde_value", palette = "viridis", title = "KDE Estimate")

tmap_mode("view")
tm_shape(kde) +
  tm_polygons(col = "kde_value", palette = "viridis", title = "KDE Estimate") +
  tm_shape(state.cs) + tm_polygons(fill="gray80", fill_alpha = 0.6)

```

#Started working on the following section in order to incorporate raster dataset in analysis, but due to resolution it failed and took a long time to run. Even tried to work on a specific region to reduce the file size, it also failed.

```{r}
# midwest <- midwest %>% terra::project(., crs(landcover))
# midwest.landcover <- landcover %>% terra::crop(.,midwest)
# midwest.landcover %>% terra::aggregate(., 20, fun = "modal") %>% mapview(.)
# 
# values <- extract(midwest.landcover, midwest)
# values
```















